The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2016-02-14 script/en/chapters/primes.tex
First version of CT Book ed 12. Most important point is the newly includes bitciphers chapter. Further things: Tried to reduce latex warnings amap. Changed wording from Sage to SageMath where visible...

\hypertarget{euclid}{}
\index{Euclid's proof by contradiction}\index{Proof by contradiction}
% \paragraph*{Euclid's proof by contradiction}
% goes as follows:
\begin{Proof}{according to Euclid (proof by contradiction)\\}
{\bf Assumption:} \quad There is a {\em finite} number of primes. \\*[4pt] {\bf
Conclusion:} \quad Then these can be listed $p_1 < p_2 < p_3 < \dots < p_n$,
where $n$ is the (finite) number of prime numbers. $p_n$ is therefore the
largest prime. Euclid now looks at the number $a = p_1 \cdot p_2 \cdots p_n +1$.
This number cannot be a prime number because it is not included in our list of
primes. It must therefore be divisible by a prime, i.e. there is a natural
number $i$ between $1$ and $n$, such that $p_i$ divides the number $a$. Of
course, $p_i$ also divides the product $a-1 = p_1 \cdot p_2 \cdots p_n$, because
$p_i$ is a factor of $a-1$. Since $ p_i $ divides the numbers $ a $ and $ a-1 $,
it also divides the difference of these numbers. Thus: $p_i$ divides  $a - (a-1)
= 1$. $p_i$ must therefore divide $1$, which is impossible. \\*[4pt] {\bf
Contradiction:} \quad Our assumption was false.

Thus there is an {\em infinite} number of primes
(Cross-reference: \hyperlink{primhfk}{overview} under \ref{s:primhfk} of 
the number of prime numbers in various intervals).
\end{Proof}

@@

\begin{Proof}{}
The theorem of Mersenne can be proved by contradiction%
\index{Proof by contradiction}.  We therefore assume that
there exists a composite natural number $ n $ (with real factorization)  
$ n=n_1 \cdot n_2 $
, with the property that $ 2^n -1 $ is a prime number.

From \begin{eqnarray*} (x^r-1)((x^r)^{s-1} + (x^r)^{s-2} + \cdots + x^r +1) & =
&  ((x^r)^s + (x^r)^{s-1} + (x^r)^{s-2} + \cdots + x^r) \\ &  & -((x^r)^{s-1} +
(x^r)^{s-2} + \cdots + x^r +1)  \\ & = & (x^r)^s -1 = x^{rs } -1,
\end{eqnarray*} we conclude \[ 2^{n_1 n_2} - 1 = (2^{n_1} -1)((2^{n_1})^{n_2 -1}
+ (2^{n_1})^{n_2 -2} + \cdots + 2^{n_1} + 1). \]
Because $ 2^n - 1 $ is a prime number, one of the above two factors on the
right-hand side must be equal to 1. This is the case if and only if $ n_1 =1 $
or $ n_2 =1$. But this contradicts our assumption. Therefore the assumption is
false. This means that there exists no composite number $ n, $ such that $ 2^n -
1 $ is a prime.
\end{Proof}

@@

% --------------------------------------------------------------------------
\subsection{\texorpdfstring{$f(n) = n^2 - 79 \cdot n + 1,601$}{f(n) = n\^{}2 - 79 ... n + 1,601}}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function delivers prime numbers for all values from $n=0$ to $n=79$
   .\footnote{%
   See chapter~\ref{primes:_Appendix_Sage-Samples},
   ``\nameref{primes:_Appendix_Sage-Samples}'' for the source code
   to compute the table using Sage\index{Sage}.}  %Pari-GP.\index{Pari-GP}}
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See chapter~\ref{primes:_Appendix_Sage-Samples}, ``\nameref{primes:_Appendix_Sage-Samples}''.).}.

@@

$M_{M_p}$ is known to be prime for p = 2, 3, 5, 7, and there has the according values:
7, 127, 2.147.483.647, 170.141.183.460.469.231.731.687.303.715.884.105.727.\footnote{%
Sage\index{Sage} calculates this with:
\begin{Verbatim}
sage: for p in (2,3,5,7): Mp=(2^p)-1; MMp=(2^Mp)-1; B=is_prime(MMp); print p, Mp, MMp, B;
....: 
2  3 7  True
3  7 127  True
5  31 2147483647  True
7  127 170141183460469231731687303715884105727  True
\end{Verbatim}
}

@@

     Arithmetic series have been known for millennia and one would think 
     they have no more secrets.
     They get more interesting again, if we impose additional constraints 
     on the serieís elements - as the prime example shows.

@@

     If someone wants to discover such sequences he should consider the
     following thought.
     The length of a sequence determines the minimal common distance
     between the single primes of the sequence. Given a sequence with 6 
     elements the distance between them has to be 30 or a multiple of 30.
     The number 30 results as the product of all primes smaller than the
     length of the sequence. So its the product of all primes smaller than 6:
     $ 6\# = 5\# = 2 * 3 * 5 = 30 $.
     It is $ 10\# = 7\# = 2 ∑ 3 ∑ 5 ∑ 7 = 210$.
     If you look for a sequence with 15 elements, then the common
     distance is at least $15\# = 13\# = 2 * 3 * 5 * 7 * 11 * 13 = 30,030$.
     \index{Prime number!k primorial} \index{Prime number!k\#}

@@

% --------------------------------------------------------------------------
% \vskip +10 pt
\newpage
\subsection{Some unproven statements / conjectures / open questions about primes}
\begin{itemize}
  \item Goldbach\footnote{%
     Christian Goldbach\index{Goldbach, Christian}, German mathematician,
     Mar 18, 1690$-$Nov 20, 1764. }
     conjectured: Every even natural number greater than 2 can
     be represented as the sum of two prime numbers.\footnote{Please compare
     chapter \ref{L-GoldbachConjecture}.}

@@

\noindent Such tables can easily be calculated using computer algebra
systems. Here is a code sample for Sage\index{Sage}\index{CAS}:

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
E = [10, 40, 56, 64, 80, 90, 112, 128, 150, 160, 192, 256, 1024, 2048]
for e in E:
    # print "2^" + str(e), "---", 1.0*(2^e)
    print "2^%4d" % e , " --- ", RR(2^e).n(24)
....:
2^  10  ---  1024.00
2^  40  ---  1.09951e12
2^  56  ---  7.20576e16
2^  64  ---  1.84467e19
2^  80  ---  1.20893e24
2^  90  ---  1.23794e27
2^ 112  ---  5.19230e33
2^ 128  ---  3.40282e38
2^ 150  ---  1.42725e45
2^ 160  ---  1.46150e48
2^ 192  ---  6.27710e57
2^ 256  ---  1.15792e77
2^1024  ---  1.79769e308
2^2048  ---  3.23170e616
\end{Verbatim}
\caption{Special values of the binary and decimal systems}
\end{sagecode}

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\newpage
\hypertarget{primes:_Appendix_Plotting-Primes-Quantity}{}
\section[Appendix: Visualization of the quantity of primes in higher ranges]
        {\sloppy Appendix: Visualization of the quantity of primes in higher ranges}
        % Wie in Fussnote eckige Klammer davor: Was in Klammern kommt in Contents
        % links, so dass dort keine Schmierzeichen mehr! 
\label{primes:_Appendix_Plotting-Primes-Quantity}
\index{Sage!Code examples}
\index{Sage}
\vskip +15pt

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\clearpage
\newpage
\hypertarget{primes:_Appendix_Sage-Samples}{}
\section{Appendix: Examples using Sage}
% \section*{Appendix A: Examples using Sage}
% \addcontentsline{toc}{section}{Appendix A: Examples using Sage}
\label{primes:_Appendix_Sage-Samples}
\index{Sage!Code examples}
\index{Sage}

\noindent Below is Sage source code related to contents of the
chapter~\ref{Label_Kapitel_Primes} (``\nameref{Label_Kapitel_Primes}'').

% ---------------------------------------------------------------------------
% \newpage
\subsection{Some basic functions about primes using Sage}
\index{Sage}

This part of the appendix contains Sage code, to perform some simple
computations about primes.%
\footnote{See the Sage documentation about Elementary number theory
          \url{http://www.sagemath.org/doc/constructions/number_theory.html}.}

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]

@@

% ---------------------------------------------------------------------------
\newpage
\subsection{Check primality of integers generated by quadratic functions}
\index{Sage}

The following Sage code verifies the primality of integers generated
by the function $f(n) = n^2 - 9n + 61$.
The code defines a function called \verb!quadratic_prime_formula()!
that takes three arguments:
\begin{itemize}
\item \verb!start! --- An integer which is the lower bound for
  integers in the sequence $\texttt{start}, \texttt{start} + 1,
  \texttt{start} + 2, \dots, \texttt{end} - 1, \texttt{end}$.

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
def quadratic_prime_formula(start, end, verbose=True):
    print "N -- N^2 - 9*N + 61"
    P = 0 # the number of primes between start and end
    for n in xrange(start, end + 1):
        X = n^2 - 9*n + 61
        if is_prime(X):
            P += 1
            if verbose:
                 print str(n) + " -- " + str(X) + " is prime"
        else:
            if verbose:
                 print str(n) + " -- " + str(X) + " is NOT prime"
    print "Number of primes: " + str(P)
    print "Percentage of primes: " + str(float((P * 100) / (end - start + 1)))
\end{Verbatim}
\caption{Verify the primality of integers generated by a quadratic function}
\end{sagecode}

@@

For larger sequences, it is impractical to print all single messages indicating
the primality of integers. In the following Sage session, only the statistics at
the end is printed (by setting the verbose parameter to false): the overall
number of primes and the percentage of primes, generated by $f(n)$ where
$0 \leq n \leq 1000$.

\begin{Verbatim}%
[fontsize=\footnotesize,fontshape=tt]
sage: quadratic_prime_formula(0, 1000, False)
N -- N^2 - 9*N + 61
Number of primes: 584
Percentage of primes: 58.3416583417
\end{Verbatim}

>>>>

\hypertarget{euclid}{}
\index{Euclid's proof by contradiction}\index{Proof by contradiction}
% \paragraph*{Euclid's proof by contradiction}
% goes as follows:
\begin{proof}{according to Euclid (proof by contradiction)\\}
{\bf Assumption:} \quad There is a {\em finite} number of primes. \\*[4pt] {\bf
Conclusion:} \quad Then these can be listed $p_1 < p_2 < p_3 < \dots < p_n$,
where $n$ is the (finite) number of prime numbers. $p_n$ is therefore the
largest prime. Euclid now looks at the number $a = p_1 \cdot p_2 \cdots p_n +1$.
This number cannot be a prime number because it is not included in our list of
primes. It must therefore be divisible by a prime, i.e. there is a natural
number $i$ between $1$ and $n$, such that $p_i$ divides the number $a$. Of
course, $p_i$ also divides the product $a-1 = p_1 \cdot p_2 \cdots p_n$, because
$p_i$ is a factor of $a-1$. Since $ p_i $ divides the numbers $ a $ and $ a-1 $,
it also divides the difference of these numbers. Thus: $p_i$ divides  $a - (a-1)
= 1$. $p_i$ must therefore divide $1$, which is impossible. \\*[4pt] {\bf
Contradiction:} \quad Our assumption was false.

Thus there is an {\em infinite} number of primes
(Cross-reference: \hyperlink{primhfk}{overview} under \ref{s:primhfk} of 
the number of prime numbers in various intervals).
\end{proof}

@@

\begin{proof}{}
The theorem of Mersenne can be proved by contradiction%
\index{Proof by contradiction}.  We therefore assume that
there exists a composite natural number $ n $ (with real factorization)  
$ n=n_1 \cdot n_2 $
, with the property that $ 2^n -1 $ is a prime number.

From \begin{eqnarray*} (x^r-1)((x^r)^{s-1} + (x^r)^{s-2} + \cdots + x^r +1) & =
&  ((x^r)^s + (x^r)^{s-1} + (x^r)^{s-2} + \cdots + x^r) \\ &  & -((x^r)^{s-1} +
(x^r)^{s-2} + \cdots + x^r +1)  \\ & = & (x^r)^s -1 = x^{rs } -1,
\end{eqnarray*} we conclude \[ 2^{n_1 n_2} - 1 = (2^{n_1} -1)((2^{n_1})^{n_2 -1}
+ (2^{n_1})^{n_2 -2} + \cdots + 2^{n_1} + 1). \]
Because $ 2^n - 1 $ is a prime number, one of the above two factors on the
right-hand side must be equal to 1. This is the case if and only if $ n_1 =1 $
or $ n_2 =1$. But this contradicts our assumption. Therefore the assumption is
false. This means that there exists no composite number $ n, $ such that $ 2^n -
1 $ is a prime.
\end{proof}

@@

% --------------------------------------------------------------------------
\subsection{\texorpdfstring{$f(n) = n^2 - 79 \cdot n + 1,601$}{f(n) = n\^{}2 - 79 ... n + 1,601}}
\label{L-Polynomfunktion02-1601}   \label{SrcPrimArith1}
   This function delivers prime numbers for all values from $n=0$ to $n=79$
   .\footnote{%
   See chapter~\ref{primes:_Appendix_Sage-Samples},
   ``\nameref{primes:_Appendix_Sage-Samples}'' for the source code
   to compute the table using SageMath\index{SageMath}.}  %Pari-GP.\index{Pari-GP}}
Unfortunately $f(80) = 1,681 = 11 \cdot 151$ is not a prime number. To this
date, no function has been found that delivers more prime numbers in a row. On
the other hand, each prime occurs twice (first in the decreasing then in the
increasing sequence), which means that the algorithm delivers a total of 40
different prime values (these are the same ones as delivered by the function
in chapter~\ref{L-Polynomfunktion01-41})\footnote{%
   Another quadratic polynom, which delivers these primes, is: 
   $f(n) = n^2 - 9 \cdot n + 61$.\\
   Among the first 1000 sequence elements more than 50\% are prime
   (See chapter~\ref{primes:_Appendix_Sage-Samples}, ``\nameref{primes:_Appendix_Sage-Samples}''.).}.

@@

$M_{M_p}$ is known to be prime for p = 2, 3, 5, 7, and there has the according values:
7, 127, 2.147.483.647, 170.141.183.460.469.231.731.687.303.715.884.105.727.\footnote{%
SageMath\index{SageMath} calculates this with:
\begin{Verbatim}
sage: for p in (2,3,5,7): Mp=(2^p)-1; MMp=(2^Mp)-1; B=is_prime(MMp); print p, Mp, MMp, B;
....: 
2  3 7  True
3  7 127  True
5  31 2147483647  True
7  127 170141183460469231731687303715884105727  True
\end{Verbatim}
}

@@

     Arithmetic series have been known for millennia and one would think 
     they have no more secrets.
     They get more interesting again, if we impose additional constraints 
     % on the serieís elements - as the prime example shows. [[to be changed because of 
     %                     Package inputenc Error: Keyboard character used is undefined
     %                     (inputenc) in inputencoding `latin1'.
     on the serie's elements - as the prime example shows.

@@

     If someone wants to discover such sequences the following thoughts may be helpful.
     The length of a sequence determines the minimal common distance
     between the single primes of the sequence. Given a sequence with 6 
     elements the distance between them has to be 30 or a multiple of 30.
     The number 30 here results from the product of all primes smaller than the
     length of the sequence (here 6):
     $ 6\# = 5\# = 2 * 3 * 5 = 30 $.
     Another example: $ 10\# = 7\# = 2 * 3 * 5 * 7 = 210$.
     If you look for a sequence with 15 elements, then the common
     distance is at least $15\# = 13\# = 2 * 3 * 5 * 7 * 11 * 13 = 30,030$.
     \index{Prime number!k primorial} \index{Prime number!k\#}

@@

% --------------------------------------------------------------------------
% \vskip +10 pt
\newpage
\subsection{Some unproven statements/ conjectures/ open questions about primes}
\begin{itemize}
  \item Goldbach\footnote{%
     Christian Goldbach\index{Goldbach, Christian}, German mathematician,
     Mar 18, 1690$-$Nov 20, 1764. }
     conjectured: Every even natural number greater than 2 can
     be represented as the sum of two prime numbers.\footnote{Please compare
     chapter \ref{L-GoldbachConjecture}.}

@@

\noindent Such tables can easily be calculated using computer algebra
systems. Here is a code sample for SageMath\index{SageMath}\index{CAS}:

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
E = [10, 40, 56, 64, 80, 90, 112, 128, 150, 160, 192, 256, 1024, 2048]
for e in E:
    # print "2^" + str(e), "---", 1.0*(2^e)
    print "2^%4d" % e , " --- ", RR(2^e).n(24)
....:
2^  10  ---  1024.00
2^  40  ---  1.09951e12
2^  56  ---  7.20576e16
2^  64  ---  1.84467e19
2^  80  ---  1.20893e24
2^  90  ---  1.23794e27
2^ 112  ---  5.19230e33
2^ 128  ---  3.40282e38
2^ 150  ---  1.42725e45
2^ 160  ---  1.46150e48
2^ 192  ---  6.27710e57
2^ 256  ---  1.15792e77
2^1024  ---  1.79769e308
2^2048  ---  3.23170e616
\end{Verbatim}
\caption{Special values of the binary and decimal systems}
\end{sagecode}

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\newpage
\hypertarget{primes:_Appendix_Plotting-Primes-Quantity}{}
\section[Appendix: Visualization of the quantity of primes in higher ranges]
        {\sloppy Appendix: Visualization of the quantity of primes in higher ranges}
        % Wie in Fussnote eckige Klammer davor: Was in Klammern kommt in Contents
        % links, so dass dort keine Schmierzeichen mehr! 
\label{primes:_Appendix_Plotting-Primes-Quantity}
\index{SageMath!Code examples}
\index{SageMath}
\vskip +15pt

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\clearpage
\newpage
\hypertarget{primes:_Appendix_Sage-Samples}{}
\section{Appendix: Examples using SageMath}
% \section*{Appendix A: Examples using SageMath}
% \addcontentsline{toc}{section}{Appendix A: Examples using SageMath}
\label{primes:_Appendix_Sage-Samples}
\index{SageMath!Code examples}
\index{SageMath}

\noindent Below is SageMath source code related to contents of the
chapter~\ref{Label_Kapitel_Primes} (``\nameref{Label_Kapitel_Primes}'').

% ---------------------------------------------------------------------------
% \newpage
\subsection{Some basic functions about primes using SageMath}
\index{SageMath}

This part of the appendix contains SageMath code, to perform some simple
computations about primes.%
\footnote{See the SageMath documentation about Elementary number theory
          \url{http://www.sagemath.org/doc/constructions/number_theory.html}.}

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]

@@

% ---------------------------------------------------------------------------
\newpage
\subsection{Check primality of integers generated by quadratic functions}
\index{SageMath}

The following SageMath code verifies the primality of integers generated
by the function $f(n) = n^2 - 9n + 61$.
The code defines a function called \verb!quadratic_prime_formula()!
that takes three arguments:
\begin{itemize}
\item \verb!start! --- An integer which is the lower bound for
  integers in the sequence $\texttt{start}, \texttt{start} + 1,
  \texttt{start} + 2, \dots, \texttt{end} - 1, \texttt{end}$.

@@

\begin{sagecode}
\begin{Verbatim}%
[fontsize=\footnotesize]
def quadratic_prime_formula(start, end, verbose=True):
    print "N -- N^2 - 9*N + 61"
    P = 0 # the number of primes between start and end
    for n in xrange(start, end + 1):
        X = n^2 - 9*n + 61
        if is_prime(X):
            P += 1
            if verbose:
                 print str(n) + " -- " + str(X) + " is prime"
        else:
            if verbose:
                 print str(n) + " -- " + str(X) + " is NOT prime"
    print "Number of primes: " + str(P)
    print "Percentage of primes: " + str(float((P * 100) / (end - start + 1)))
\end{Verbatim}
\caption{Verify the primality of integers generated by a quadratic function}
\end{sagecode}

@@

For larger sequences, it is impractical to print all single messages indicating
the primality of integers. In the following SageMath session, only the statistics at
the end is printed (by setting the verbose parameter to false): the overall
number of primes and the percentage of primes, generated by $f(n)$ where
$0 \leq n \leq 1000$.

\begin{Verbatim}%
[fontsize=\footnotesize]
sage: quadratic_prime_formula(0, 1000, False)
N -- N^2 - 9*N + 61
Number of primes: 584
Percentage of primes: 58.3416583417
\end{Verbatim}
